Efficient pedigree recording for fast population genetics simulation

### 2017-12-09 forwards_paper.tex
added other simplify complexity info

\emph{Note:} author order not determined

@@

%%%%%%%%%%%%%%
\subsection*{Sequential simplification and prior history}

@@

Consider a simulation of a Wright-Fisher population of $2N$ haploid individuals
using Algorithm~\algref{W} for $T$ generations,
with simplify interval $s = 1$ so that redundant information is removed after every generation.
Since every chromosome inherits material from both parents,
without simplification this would produce tables of
$2NT$ nodes and
$4NT$ edges.
Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability,
($T \sim 20N$, say).
After simplification, we are left with the tree sequence describing the history
of only the current generation of $2N$ individuals.
This tree sequence has $4N-2$ edges to describe the leftmost tree;
and each time the marginal tree changes along the sequence,
four edges end and four new edges begin 
(except for changes affecting the root, which require fewer; see \citet{kelleher2016efficient}).
Coalescent theory tells us that
the expected total length of the edges of a marginal tree is approximately $4N\log(2N)$,
which is also equal to the mean number of ancestral recombination events that occur on a branch of the marginal tree,
since we have taken one crossover per individual per generation.
Not all such recombinations actually change the tree topology,
so four times this gives us an upper bound on the expected number of additional edges.
Similarly, not every new edge derives from a never-before-seen node,
but the number of nodes is at most equal to the number of edges plus the sample size.
With $T=20N$, this reduces the initial storage of $120 N^2$ items to 
% 4N - 2 + 2N + 4N log(2N) * 4 * 2 = 6N + 32 N log(2N)
$6N + 32 N \log(2N)$;
at $N=10^4$, a factor of 3,700.

To get an idea of how required space depends on the length of time between simplification steps,
suppose instead that we have run a simulation of $2N$ individuals for $T$ generations,
begun with no prior history.
How many of the resulting $4NT$ edges are required after simplification?
As above, the expected number of edges is bounded by $2N-2$ plus four times the mean marginal tree length.
Roughly, the length of time for which a marginal tree has $k$ tips is
$2N/k(k-1) = 2N(1/(k-1) - 1/k)$,
and so the time to go from $N$ to $n$ tips is $2N(1/(n-1) - 1/(N-1))$.
This implies that after running for time $T$ we expect to
have around $n(T)$ roots, where $n(T) \approx 1 + 2N/(T+2)$.
The total tree length over this time is
$2 N \sum_{k=n(T)+1}^{N-1} 1/k$, which
% is approximately
% \begin{align*}
%     2 N \log\left( \frac{N}{1 + 2 N/(T+2)} \right) .
% \end{align*}
leads to an upper bound on the number of edges of
\begin{align}
    \label{eqn:edge_bound}
    2 N \left( 1 + 4 \log\left( \frac{NT}{T + 2 N} \right)\right) .
\end{align}
This holds up quite well in practice,
as shown in Figure~\ref{fig:num_edges}.
% by comparison to
% the number of edges actually required for 50 independent Wright--Fisher simulations,
% as a function of time (obtained from Algorithm~\algref{W} with $s=1$)
% is shown in Figure~\ref{fig:num_edges}, compared to this prediction.

\begin{figure}
    \begin{center}
        \includegraphics{sims/num_edges_50}
    \end{center}
    \caption{
        Number of edges in the simplified tree sequence
        for 50 Wright--Fisher simulations begun with no prior history,
        as a function of number of generations.
        Each line is one simulation, the black line gives the average,
        and the dotted line is the upper bound of equation \eqref{eqn:edge_bound}.
        \label{fig:num_edges}
    }
\end{figure}

If we were to add mutations in forwards time
under the infinite-sites model with total mutation rate per generation $\mu$,
there would also be around $\mu 2NT$ mutations (and the same number of sites).
Since mutations fall as recombinations do on the marginal trees,
adding them after simplification results in only around $\mu 4 N \log(2N)$ mutations,
and the same savings as we obtained from edges above,
even without considering the 
computational burden of propagating mutations forwards across generations.

Our method stores genealogies, and so records substantially more information
than would a method only recording genotypes.
However, since the simplification algorithm requires computational effort,
it is informative to compare computational complexity of the algorithm
to one that propagates neutral genotypes.
A typical individual differs at around $2 N \mu$ sites from the population's consensus sequence,
so propagating these to offspring by simple copying will take $4 N^2 \mu$ operations per generation.
On the other hand, 
in our scheme we must store $13N$ quantities per generation (two edges and one node per individual).
The simplification algorithm is linear in the number of edges of the input tree sequence,
and so multiplies this by a constant factor.
This implies that propagating neutral genotypes for $T$ generations has time complexity $O(\mu T N^2)$,
while our implementation is only $O(T N \log(2N))$.

@@

The tree sequences produced by default by this method
are very compact, storing genotype \emph{and} genealogical information
in a small fraction of the space taken by a compressed VCF file.
The format is also highly efficient to process,
leading to advantages for downstream analysis as well.
This is because many algorithms to compute statistics of interest for population genetics
are naturally expressed in terms of tree topologies,
and so can be quickly computed from the trees underlying the tree sequence format.
For example, pairwise nucleotide diversity $\pi$, is the average density of
differences between sequences in the sample.
To compute this directly from sequence data at $m$ sites in $n$ samples
requires computing allele frequencies, taking $O(nm)$ operations.
By using the locations of the mutations on the marginal trees,
and the fact that these are correlated,
sequential tree algorithms similar to those in~\citep{kelleher2016efficient}
can do this in roughly $O(n + m \log n)$ operations.
The \msprime{} API provides a method to compute $\pi$ among arbitrary subsets of the
samples in a tree sequence, which took about 1.2 seconds
applied to the example tree sequence above with 1.1 million mutations
in 200 megabases for 500,000 samples.

>>>>

@@

%%%%%%%%%%%%%%
\subsubsection*{Sequential simplification and prior history}

@@

\plr{Refer to simulation results here.}

%% Memory reduction thanks to simplification
Consider a simulation of a Wright-Fisher population of $2N$ haploid individuals
using Algorithm~\algref{W} for $T$ generations,
with simplify interval $s = 1$ so that redundant information is removed after every generation.
Since every chromosome inherits material from both parents,
without simplification this would produce tables of
$2NT$ nodes and
$4NT$ edges.
Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability,
($T \sim 20N$, say).
After simplification, we are left with the tree sequence describing the history
of only the current generation of $2N$ individuals.
This tree sequence has $4N-2$ edges to describe the leftmost tree;
and each time the marginal tree changes along the sequence,
four edges end and four new edges begin 
(except for changes affecting the root, which require fewer; see \citet{kelleher2016efficient}).
Coalescent theory tells us that
the expected total length of the edges of a marginal tree is approximately $4N\log(2N)$,
which is also equal to the mean number of ancestral recombination events that occur on a branch of the marginal tree,
since we have taken one crossover per individual per generation.
Not all such recombinations actually change the tree topology,
so four times this gives us an upper bound on the expected number of additional edges.
Similarly, not every new edge derives from a never-before-seen node,
but the number of nodes is at most equal to the number of edges plus the sample size.
With $T=20N$, this reduces the initial storage of $120 N^2$ items to 
% 4N - 2 + 2N + 4N log(2N) * 4 * 2 = 6N + 32 N log(2N)
$6N + 32 N \log(2N)$;
at $N=10^4$, a factor of 3,700.

%% Progressive memory reduction as a function of T
To get an idea of how required space depends on the length of time between simplification steps,
suppose instead that we have run a simulation of $2N$ individuals for $T$ generations,
begun with no prior history.
How many of the resulting $4NT$ edges are required after simplification?
As above, the expected number of edges is bounded by $2N-2$ plus four times the mean marginal tree length.
Roughly, the length of time for which a marginal tree has $k$ tips is
$2N/k(k-1) = 2N(1/(k-1) - 1/k)$,
and so the time to go from $N$ to $n$ tips is $2N(1/(n-1) - 1/(N-1))$.
This implies that after running for time $T$ we expect to
have around $n(T)$ roots, where $n(T) \approx 1 + 2N/(T+2)$.
The total tree length over this time is
$2 N \sum_{k=n(T)+1}^{N-1} 1/k$, which
% is approximately
% \begin{align*}
%     2 N \log\left( \frac{N}{1 + 2 N/(T+2)} \right) .
% \end{align*}
leads to an upper bound on the number of edges of
\begin{align}
    \label{eqn:edge_bound}
    2 N \left( 1 + 4 \log\left( \frac{NT}{T + 2 N} \right)\right) .
\end{align}
This holds up quite well in practice,
as shown in Figure~\ref{fig:simplify_complexity}.
% by comparison to
% the number of edges actually required for 50 independent Wright--Fisher simulations,
% as a function of time (obtained from Algorithm~\algref{W} with $s=1$)
% is shown in Figure~\ref{fig:simplify_complexity}, compared to this prediction.

\begin{figure}
    \begin{center}
        \includegraphics{sims/num_edges_50}
        \includegraphics{sims/simplify_timings}
    \end{center}
    \caption{
        \textbf{(A)}
        Number of edges in the simplified tree sequence
        for 50 Wright--Fisher simulations begun with no prior history,
        as a function of number of generations.
        Each line is one simulation, the black line gives the average,
        and the dotted line is the upper bound of equation \eqref{eqn:edge_bound}.
        \plr{Propose to add other figures as subfigures here:}
        \textbf{(B)}
        Time required to simplify the first $k$ edges
        of \plr{what tree sequence?},
        plotted against $k$.
        \textbf{(C)}
        Time required to simplify
        \plr{what tree sequence?}
        down to a sample of size $n$,
        plotted against $n$
        (note the log scale;
        the time scales logarithmically with $n$).
        \label{fig:simplify_complexity}
    }
\end{figure}

%% Above was edges; same holds for memory usage savings due to mutations
If we were to add mutations in forwards time
under the infinite-sites model with total mutation rate per generation $\mu$,
there would also be around $\mu 2NT$ mutations (and the same number of sites).
Since mutations fall as recombinations do on the marginal trees,
adding them after simplification results in only around $\mu 4 N \log(2N)$ mutations,
and the same savings as we obtained from edges above,
even without considering the 
computational burden of propagating mutations forwards across generations.

%% simplify run time
Above we have described memory requirements;
how does the computation time required for simplification scale?
Simply because it must process each edge,
the simplification algorithm is linear in the number of edges of the input tree sequence.
This is seen in Figure~\ref{fig:simplify_complexity}X,
where we show shows the time required to simplify increasingly large subsets of a large tree sequence.
When simplifying the result of a forwards-time sequence, the number of edges is the main contributing factor.
Suppose on the other hand we want to 
simplify an already-minimal but large tree sequence with $N$ nodes
to a subsample of size $n$.
How does the required time scale with $n$?
In this case, the computation is dominated by the size of the output tree sequence,
which grows with $\log(n)$.
this is also seen in Figure~\ref{fig:simplify_complexity}X,
showing the time required to simplify progressively larger subsets of a single large tree sequence.

Our method stores genealogies, and so records substantially more information
than would a method only recording genotypes.
However, since the simplification algorithm requires computational effort,
it is informative to compare computational complexity of the algorithm
to one that propagates neutral genotypes.
A typical individual differs at around $2 N \mu$ sites from the population's consensus sequence,
so propagating these to offspring by simple copying will take $4 N^2 \mu$ operations per generation.
On the other hand, 
in our scheme we must store $13N$ quantities per generation (two edges and one node per individual).
\plr{XXX}
The simplification algorithm is linear in the number of edges of the input tree sequence
(because it must process each edge),
and so multiplies this by a constant factor.
This implies that propagating neutral genotypes for $T$ generations has time complexity $O(\mu T N^2)$,
while our implementation is only $O(T N \log(2N))$.

@@

The tree sequences produced by default by this method
are very compact, storing genotype \emph{and} genealogical information
in a small fraction of the space taken by a compressed VCF file.
The format is also highly efficient to process,
leading to advantages for downstream analysis as well.
This is because many algorithms to compute statistics of interest for population genetics
are naturally expressed in terms of tree topologies,
and so can be quickly computed from the trees underlying the tree sequence format.
For example, pairwise nucleotide diversity $\pi$, is the average density of
differences between sequences in the sample.
To compute this directly from sequence data at $m$ sites in $n$ samples
requires computing allele frequencies, taking $O(nm)$ operations.
By using the locations of the mutations on the marginal trees,
and the fact that these are correlated,
sequential tree algorithms similar to those in~\citep{kelleher2016efficient}
can do this in roughly $O(n + m \log n)$ operations.
The \msprime{} API provides a method to compute $\pi$ among arbitrary subsets of the
samples in a tree sequence, which took about 1.2 seconds
applied to the example tree sequence above with 1.1 million mutations
in 200 megabases for 500,000 samples.
Calculating $\pi$ from the numeric genotype matrix using numpy took \plr{HOW MANY} seconds.
